Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated

### 2015-05-01 manuscript.Rmd
Lots of progress on manuscript, most references done, all (?) stats added. captions to be done and intro + abstract to be rewritten

---
title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
  atmospheric [$CO_2$] but tracks water availability
author: "Remko Duursma"
date: "April 24, 2015"
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: global-change-biology.csl
bibliography: references.bib
---

```{r echo=FALSE, message=FALSE}
library(knitr)
load("cache/lai_workspace.RData")
source("load.R")
library(broom)
options(scipen = 1, digits = 2)

@@

@@

@@

*Keywords*: leaf area index, FACE, phenology, drought, litter production

@@

Direct effects of $eC_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012]. Although numerous studies have been conducted on the response of physiology and plant growth to $eC_a$, very few of these inform the response of steady-state *L* to $eC_a$, because it requires a sufficiently mature forest to be exposed to free air $CO_2$ enrichment (FACE) [@Norby2011]. Increased carbon availability through enhanced photosynthesis under $eC_a$ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to $eC_a$ *L* [@Norby2011], although positive responses to $eC_a$ in terms of productivity were observed at all six sites. Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to $eC_a$, leading to the prediction that stands with low *L* will show a larger response to $eC_a$. Because diminished returns in increasing LAI in response to $eC_a$ when LAI is larger [@Norby2005].

In ecosystems where water availability limits productivity, we could expect $eC_a$ to further affect *L* [@Donohue2009], since $eC_a$ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance. **Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.** In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising $C_a$. Based on the same water-savings principle, we predict that *L* should also increase under $eC_a$ in water-limited forests.

Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests (e.g. [@Richardson2012; @Savoy2015]) whereas *L* responses to fluctuations in water availability are qualitatively well-known (**more cites**), but poorly quantified. The *L* of broad-leaved evergreen trees that dominate forests in the southern hemisphere show particular sensitivity to water-availability [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009].The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability. Nevertheless, few datasets are currently available at sufficiently high temporal resolution to allow modelling of water-driven changes in phenology [@Misson2011].

We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to $eC_a$, the EucFACE experiment. The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that $eC_a$ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated $CO_2$. In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to $eC_a$ [@Norby2011]. Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

The soil at the site is a loamy sand in the Clarendon Formation (Barton et al., 2010; Isbell, 2002), a slightly-acid, low-organic C, well-drained soil. The soil surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 – 12.9 m below the soil surface. The woodland ranges in density from 600-1000 trees $ha^{-1}$. The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m$^2$ ha$^{-1}$, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E. amplifolia* Naudin. The understory stratum is composed mainly of grass species dominated by *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted $CO_2$ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009]. The vegetation within three randomly selected rings was exposed to a $CO_2$ mole fraction ($C_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the $CO_2$ treatment rings in infrastructure and instrumentation but received ambient $C_a$, and are used as control plots. The target diurnal $C_a$ for the rings exposed to $eC_a$ was 150 μmol $CO_2$ mol$^{-1}$ above ambient $C_a$ in full operation mode. However, $C_a$ was increased gradually for six months at a rate of ca. 30 μmol $CO_2$ mol$^{-1}$ month$^{-1}$ until full operation mode was reached (Figure 1a).

Air temperature was measured on top of the central tower of each ring (23.5 m height) with a HUMICAP sensor (HMP 155 Vaisala, Vantaa, Finland). Total precipitation was monitored with three automated tipping buckets (Tipping Bucket Rain gauge TB4, Hydrological Services Pty Ltd, Liverpool, NSW, Australia) located at the top of the central tower at 23.5 m height, in three of the rings. Total precipitation and mean temperature were logged every 15 minutes onto CR3000 data loggers. Soil volumetric water content was measured at two different locations in each ring with two sets of theta probes (ThetaProbe ML2x, Delta-T Devices Ltd, Cambridge, UK) installed at 5, 30, 75 cm depth and at the depth at which a hard pan with higher clay content was present (30-50cm depth depending on the ring).

@@

## Estimation of leaf area index from digital photography
Leaf area index (*L*) was estimated from zenith canopy photographs following [@Macfarlane2007]. From October 2012 to October 2013, we took ca. monthly canopy photographs in each FACE ring. Photographs were taken in the morning on cloudy days. We took one photograph at each point of a 4 x 4 m grid established within each ring, unless obstructed by trees, yielding 20-23 photos per ring. Sampling points too close to either the vertical vent pipes (edges of the ring) or the central scaffolding tower were excluded. The camera was aligned with the north-south axis and set at 1.25 m height.  Camera lens was pointed directly upwards and levelled using a three axis bubble level. We used a Nikon D90 camera with a Nikon AF 50 mm lens (Nikon Corporation, Tokyo, Japan). To measure cover, the camera was set to automatic exposure, Aperture-Priority and automatic autofocus mode and F10 lens. Following recommendations by Jonckheere et al. [-@Jonckheere2005], we used the automated Ridler algorithm to threshold the canopy photos. We then calculated canopy transmittance (τ_p) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold. *L* from canopy photos (Lp) was calculated with a simple Beer-Lambert law (Eq. 1).

where τ_p canopy transmission estimated from canopy photos, and we assumed that the extinction coefficient k = 0.5, following (Macfarlane et al., 2007) recommendation for eucalypt forests.

## *L* from diffuse transmittance
In each of the six FACE rings, 3 photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height similar to the approach used by Norby et al. 2003. Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 Oct. 2012, and automatically uploaded every day to a data management system. At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height. All PAR data were averaged in 30 min time steps. At a nearby site (within 2 km), we measured the diffuse fraction (Fdiff) of PAR at 30 min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending above the canopy. For determination of *L*, we used only highly diffuse conditions, where Fdiff > 0.98. We further trimmed the data by excluding low solar elevation (< 10°), because, for unknown reasons, measured canopy transmittance was more variable at these low solar angles. Finally canopy transmittance was calculated for every 30-min time step as,

@@

We then estimated leaf area index (*L*) from $\tau_d$ ($L_\tau$), following (Campbell and Norman, 2000). We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space. We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the *L* estimate to leaf litter production as further detailed below. Diffuse transmittance of a canopy is a function of the *L* by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).

@@

where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam (Campbell and Norman, 2000). Because we assumed a spherical leaf angle distribution, G(θ) = 0.5 (Stenberg, 2006). We solved numerically for *L* in Eq. 3, using a smooth integration over $\theta$, yielding *L*τ.

To overcome violations in the assumptions underlying Eq. 3, we calibrated *L* against litter production. We estimated *L* from $\tau_d$ ($L_\tau$) and calibrating against cumulative litter production during a drought period in 2013 (indicated in Fig. 1). We selected a period of sustained drought in 2013 (14 July 2013 – 12 Nov 2013), during which no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During this period, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station in operation since 1881 (compared to a long-term average of 188 mm for this period). A total of n=11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period. In theory, cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) during that period should thus match the observed decrease in *L* estimated from canopy transmittance. We calibrated equation (4),

@@

## Data analysis
We used generalized additive models (gam) for all analyses of *L* and $\tau_d$ timeseries (Wood, 2006), using the *mgcv* package in R [@RCT2015]. For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics. In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig. 7). We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson (2014). For comparison between *L* from photos and PAR transmittance, we used simple linear regression. To test for effects of $C_a$ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and $C_a$ treatment and date as a factor.

To assess effects of $eC_a$ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows. The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the $eC_a$ treatment across the FACE rings (Fig. 3). Because the EucFACE woodland is well established, the differences in BA are not due to $C_a$ treatment effects, but instead represent natural variation within the woodland. We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig. 3, to yield a ‘*L* anomaly’. This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.

@@

Canopy transmittance ($\tau_d$) varied between 0.10 and 0.36, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig. 1).  However, temporal trends were very similar between the rings. Since the understorey comprises mostly grasses and forbs, and the PAR sensors were installed at 1 m height, this variation is due to the overstorey trees only.

An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept. 2014 (for description of a similar outbreak in *Eucalyptus moluccana*, see (Hall et al., 2015)). This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig. 1). However, it recovered rapidly after rainfall in December 2014, resulting in a lower τ_d than before the insect attack.

We compared calibrated $L_\tau$ to an independent estimate of *L* from canopy photos ($L_p$), which were collected nine times between Oct. 2012 and Oct. 2013. Without further calibration, the two estimates were reasonably well correlated, though the canopy photos gave slightly higher estimates of *L* (Fig. 2, R2=0.71, RMSE=0.12). This is encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq. 3), weighted towards the zenith.

We found no differences in the *L* anomaly between the $C_a$ treatments (Fig. 4) (p > 0.05), since the 95% confidence intervals of the generalized additive model overlapped. A linear mixed-effects model with $C_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of $C_a$ on the *L* anomaly (P > 0.2). What difference between $C_a$ treatments could we have reasonably detected? The average half-width of the confidence interval on the *L* anomaly (Fig. 4) was 0.047 m$^2$ m$^{-2}$ for $aC_a$, and 0.042 m$^2$ m$^{-2}$ for the $eC_a$ treatment. This means that, provided the within-treatment variance stays the same, we could detect a difference of 0.089 m$^2$ m$^{-2}$ (the sum of the half-width CI’s), or about a change of 6% over the mean *L* of the EucFACE (1.39 m$^2$ m$^{-2}$).

The dynamics of $L_\tau$ were compared with litter fall, by estimating the absolute change in $L_\tau$ for every FACE ring, for every period when litter was collected. We did this by fitting a generalized additive model to $L_\tau$ for each ring separately (similar to shown for $\tau_d$ in Fig. 1), and calculating the change in $L_\tau$ for each period ($\Delta L_\tau$). Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig. 5). During periods of net leaf loss, we found a significant correlation (with a slope of XX ± se, P < 0.001?, df or n?) close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing. During periods of net leaf growth, we found a significant correlation (slope, P, df…) between $\Delta L_\tau$ and litter production (Fig. 5).

We found a significant correlation among FACE rings between litter production (averaged over the entire period) and mean *L*τ during the same period (Fig. 6a), which is expected if leaf lifespan is relatively constant. This allowed us estimate mean leaf lifespan from *L*τ and litter production as the ratio of *L*τ to annual litter production, yielding a mean lifespan of 0.95 yr, with no significant difference (P? F? df?) between $C_a$ treatments (Fig. 6b).

>>>>

---
title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
  atmospheric [$CO_2$] but tracks water availability
author: ""
date: ""
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: global-change-biology.csl
bibliography: references.bib
---

```{r echo=FALSE, message=FALSE}
library(knitr)
load("cache/lai_workspace.RData")
source("load.R")

library(broom)
library(lme4)
library(lmerTest)
library(car)

# Number of digits in printing of numbers
options(scipen = 1, digits = 2)

@@

# Comparison flat canopy photo with transmittance
lmphoto <- lm(LAI.PAR.mean ~ LAI, data=flatcan_byring)

# Significance CO2 on LAI anomaly
df$Time <- as.factor(df$Date - min(df$Date))
m3 <- lmer(LAIanomaly ~ treatment*Time + (1|Ring), data=df)
LanomAOV <- Anova(m3)

# Average CI width in Fig 4.
fit <- smoothplot(Date, LAIanomaly, g=treatment, R="Ring", 
                  data=facegap_cloudy_byring, plot=FALSE, kgam=.gambase)
pa <- predict(fit$ambient, data.frame(X=as.numeric(facegap_cloudy_byring$Date)), se.fit=TRUE)
pe <- predict(fit$elevated, data.frame(X=as.numeric(facegap_cloudy_byring$Date)), se.fit=TRUE)
LanomCIamb <- mean(pa$se.fit)*2
LanomCIele <- mean(pe$se.fit)*2

# Detectable Ca effect on LAI
Ldetect <- 100 * (LanomCIamb + LanomCIele) / mean(ba$LAI.mean)

# Litter - dLAI correlations Figure 4
dl <-dLAIlitter
dl$Y <- with(dl, dLAI * 30.5/ndays)
dl$X <- with(dl, dLAIlitter.mean * 30.5/ndays)
dlup <- dl[dl$LAIchange == "increasing",]
dldown <- dl[dl$LAIchange == "decreasing",]
lmlitup <- lm(Y ~ X, data=dlup)
lmlitdown <- lm(Y ~ X, data=dldown)

# Leaf lifespan
LLtest <- tidy(t.test(LL ~ treatment,data=ba))

@@

\
$\mathrm{Remko~A.~Duursma}^1$, $\mathrm{Teresa~E.~Gimeno}^{1,2}$, $\mathrm{Matthias~M.~Boer}^1$, $\mathrm{Kristine~Y.~Crous}^1$, $\mathrm{Mark~G.~Tjoelker}^1$, $\mathrm{David~S.~Ellsworth}^1$

\
$~^1$ Hawkesbury Institute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW, Australia

\
$~^2$ INRA, UMR 1391 ISPA, 33140 Villenave d'Ornon, France

@@

\
*Keywords*: leaf area index, FACE, phenology, drought, litter production

@@

\
Direct effects of $eC_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012]. Although numerous studies have been conducted on the response of physiology and plant growth to $eC_a$, very few of these inform the response of steady-state *L* to $eC_a$, because it requires a sufficiently mature forest to be exposed to free air $CO_2$ enrichment (FACE) [@Norby2011]. Increased carbon availability through enhanced photosynthesis under $eC_a$ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to $eC_a$ *L* [@Norby2011], although positive responses to $eC_a$ in terms of productivity were observed at all six sites. Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to $eC_a$, leading to the prediction that stands with low *L* will show a larger response to $eC_a$. Because diminished returns in increasing *L* in response to $eC_a$ when LAI is larger [@Norby2005].

\
In ecosystems where water availability limits productivity, we could expect $eC_a$ to further affect *L* [@Donohue2009], since $eC_a$ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance. **Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.** In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising $C_a$. Based on the same water-savings principle, we predict that *L* should also increase under $eC_a$ in water-limited forests.

\
Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests (e.g. [@Richardson2012; @Savoy2015]) whereas *L* responses to fluctuations in water availability are qualitatively well-known (**more cites**), but poorly quantified. The *L* of broad-leaved evergreen trees that dominate forests in the southern hemisphere show particular sensitivity to water-availability [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009].The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability. Nevertheless, few datasets are currently available at sufficiently high temporal resolution to allow modelling of water-driven changes in phenology [@Misson2011].

\
We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to $eC_a$, the EucFACE experiment. The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that $eC_a$ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated $CO_2$. In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to $eC_a$ [@Norby2011]. Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

\
The soil at the site is a loamy sand in the Clarendon Formation (Barton et al., 2010; Isbell, 2002), a slightly-acid, low-organic C, well-drained soil. The soil surface is a loamy sand with > 75% sand content, and from ca. 50 to 300 cm depth, soils are a sandy clay loam with > 30% silt and clay. The soil is nutrient poor, and a strong limitation of P on tree growth at the site has recently been demonstrated [@Crous2015]. Groundwater depth is 12.6 – 12.9 m below the soil surface. The woodland ranges in density from 600-1000 trees $ha^{-1}$. The vegetation consists of an open woodland (basal area before treatment, BA = `r mean(ba$BA)`, SE = `r sd(ba$BA)/sqrt(6)` m$^2$ ha$^{-1}$, n = 6 plots) with a single major canopy-forming tree (*Eucalyptus tereticornis* Sm.) with a minor component of *E. amplifolia* Naudin. The understory stratum is composed mainly of grass species dominated by *Microlaena stipoides* (Labill.) R.Br., with lower densities of forbs of *Commelina cyanea* R.Br., *Lobelia purpurascens* R.Br., *Dichondra repens* J.R.Forst. & G.Forst and occasional shrubs of *Breynia oblongifolia* (Mull.Arg.) Mull.Arg. and *Bursaria spinosa* Cav.

\
Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum [@Hendrey1999] through which a computer-controlled system modulates release of diluted $CO_2$ into the vegetation according to a proportional-integral-differential type of algorithm [@Lewin2009]. The vegetation within three randomly selected rings was exposed to a $CO_2$ mole fraction ($C_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the $CO_2$ treatment rings in infrastructure and instrumentation but received ambient $C_a$, and are used as control plots. The target diurnal $C_a$ for the rings exposed to $eC_a$ was 150 μmol $CO_2$ mol$^{-1}$ above ambient $C_a$ in full operation mode. However, $C_a$ was increased gradually for six months at a rate of ca. 30 μmol $CO_2$ mol$^{-1}$ month$^{-1}$ until full operation mode was reached (Fig. 1a).

\
Air temperature was measured on top of the central tower of each ring (23.5 m height) with a HUMICAP sensor (HMP 155 Vaisala, Vantaa, Finland). Total precipitation was monitored with three automated tipping buckets (Tipping Bucket Rain gauge TB4, Hydrological Services Pty Ltd, Liverpool, NSW, Australia) located at the top of the central tower at 23.5 m height, in three of the rings. Total precipitation and mean temperature were logged every 15 minutes onto CR3000 data loggers. Soil volumetric water content was measured at two different locations in each ring with two sets of theta probes (ThetaProbe ML2x, Delta-T Devices Ltd, Cambridge, UK) installed at 5, 30, 75 cm depth and at the depth at which a hard pan with higher clay content was present (30-50cm depth depending on the ring).

@@

## *L* from diffuse transmittance
In each of the six FACE rings, 3 photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height similar to the approach used by Norby et al. [-@Norby2003]. Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 Oct. 2012, and automatically uploaded every day to a data management system. At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height. All PAR data were averaged in 30 min time steps. At a nearby site (within 2 km), we measured the diffuse fraction (Fdiff) of PAR at 30 min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending above the canopy. For determination of *L*, we used only highly diffuse conditions, where Fdiff > 0.98. We further trimmed the data by excluding low solar elevation (< 10°), because, for unknown reasons, measured canopy transmittance was more variable at these low solar angles. Finally canopy transmittance was calculated for every 30-min time step as,

@@

\
We then estimated leaf area index (*L*) from $\tau_d$ ($L_\tau$), following Campbell and Norman [-@Campbell2000]. We assumed that leaves had a spherical leaf angle distribution, and leaves were randomly distributed in space. We recognize that these assumptions are very likely violated, but we account for this bias by calibrating the *L* estimate to leaf litter production as further detailed below. Diffuse transmittance of a canopy is a function of the *L* by integrating the Beer-Lambert law across the entire sky hemisphere, and accounting for differences in extinction coefficient with different viewing angles (because of leaf angle and longer path length through the canopy at low solar angles).

@@

where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [-@Campbell2000]. Because we assumed a spherical leaf angle distribution, G(θ) = 0.5 [@Stenberg2006]. We solved numerically for *L* in Eq. 3, using a smooth integration over $\theta$, yielding $L_\tau$.

\
To overcome violations in the assumptions underlying Eq. 3, we calibrated *L* against litter production. We estimated *L* from $\tau_d$ ($L_\tau$) and calibrating against cumulative litter production during a drought period in 2013 (indicated in Fig. 1). We selected a period of sustained drought in 2013 (14 July 2013 – 12 Nov 2013), during which no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During this period, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station in operation since 1881 (compared to a long-term average of 188 mm for this period). A total of n=11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period. In theory, cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) during that period should thus match the observed decrease in *L* estimated from canopy transmittance. We calibrated Eq. 4,

@@

## Estimation of leaf area index from digital photography

We took advantage of the detailed timeseries of *L* estimates from canopy transmittance to test a simple photography-based method proposed by Macfarlane et al. [-@Macfarlane2007], which is extensively applied to *Eucalyptus*-dominated forests. From October 2012 to October 2013, we took ca. monthly canopy photographs in each FACE ring. Photographs were taken in the morning on cloudy days. We took one photograph at each point of a 4 x 4 m grid established within each ring, unless obstructed by trees, yielding 20-23 photos per ring. Sampling points too close to either the vertical vent pipes (edges of the ring) or the central scaffolding tower were excluded. The camera was aligned with the north-south axis and set at 1.25 m height.  Camera lens was pointed directly upwards and levelled using a three axis bubble level. We used a Nikon D90 camera with a Nikon AF 50 mm lens (Nikon Corporation, Tokyo, Japan). To measure cover, the camera was set to automatic exposure, Aperture-Priority and automatic autofocus mode and F10 lens. Following recommendations by Jonckheere et al. [-@Jonckheere2005], we used the automated Ridler algorithm to threshold the canopy photos. We then calculated canopy transmittance ($\tau_p$) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold. *L* from canopy photos ($L_p$) was calculated with a simple Beer-Lambert law (Eq. 1).

where $\tau_p$ canopy transmission estimated from canopy photos, and we assumed that the extinction coefficient k = 0.5, following Macfarlane et al. [-@Macfarlane2007] recommendation for eucalypt forests.

\
We compared $L_\tau$ (after calibration to litter production) to the independent estimate of *L* from canopy photos ($L_p$). Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)` higher than $L_p$ (Fig. SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`). This is encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq. 3), weighted towards the zenith.

## Data analysis
We used generalized additive models (gam) for all analyses of *L* and $\tau_d$ timeseries [@Wood2006], using the *mgcv* package in R [@RCT2015]. For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics. In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig. 7). We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson [@Simpson2014]. For comparison between *L* from photos and PAR transmittance, we used simple linear regression. To test for effects of $C_a$ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and $C_a$ treatment and date as a factor.

\
To assess effects of $eC_a$ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows. The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the $eC_a$ treatment across the FACE rings (Fig. 3). Because the EucFACE woodland is well established, the differences in BA are not due to $C_a$ treatment effects, but instead represent natural variation within the woodland. We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig. 3, to yield a ‘*L* anomaly’. This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.

@@

Canopy transmittance ($\tau_d$) varied between `r min(df$Gapfraction.mean)` and `r max(df$Gapfraction.mean)`, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig. 1).  However, temporal trends were very similar between the rings. Since the understorey comprises mostly grasses and forbs, and the PAR sensors were installed at 1 m height, this variation is due to the overstorey trees only.

\
An outbreak of psyllids (*Cardiaspina* sp.), a tiny sap-feeding insect, resulted in rapid loss of canopy leaf area in one of the rings, starting in Sept. 2014 (for description of a similar outbreak in *Eucalyptus moluccana*, see [@Hall2015]). This rapid decrease in *L* was evident in a quick increase in $\tau_d$ (Fig. 1). However, it recovered rapidly after rainfall in December 2014, resulting in a lower $\tau_d$ than before the insect attack.

\
We found no differences in the *L* anomaly between the $C_a$ treatments, as shown by consistent overlap of the the 95% confidence intervals of the generalized additive model  (thus, p > 0.05 at any timepoint, Fig. 3). A brief exception was the period where one of the rings was affected by an insect outbreak (see above). 
A linear mixed-effects model with $C_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of $C_a$ on the *L* anomaly (p = `r LanomAOV[[3]][1]`). What difference between $C_a$ treatments could we have reasonably detected? The average half-width of the confidence interval on the *L* anomaly (Fig. 4) was `r LanomCIamb` $m^2$ $m^{-2}$ for $aC_a$, and `r LanomCIele` $m^2$ $m^{-2}$ for the $eC_a$ treatment. This means that, provided the within-treatment variance stays the same, we could detect a difference of `r LanomCIamb + LanomCIele` $m^2$ $m^{-2}$ (the sum of the half-width CI’s), or about a change of `r Ldetect`% over the mean *L* of the EucFACE (`r mean(ba$LAI.mean)` $m^2$ $m^{-2}$).

\
The dynamics of $L_\tau$ were compared with litter fall, by estimating the absolute change in $L_\tau$ for every FACE ring, for every period when litter was collected. We did this by fitting a generalized additive model to $L_\tau$ for each ring separately (similar to shown for $\tau_d$ in Fig. 1), and calculating the change in $L_\tau$ for each period ($\Delta L_\tau$). Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig. 4). During periods of net leaf loss, we found a significant correlation (with a slope of `r tidy(lmlitdown)$estimate[2]`, SE = `r tidy(lmlitdown)$estimate[2]`, n = `glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing. During periods of net leaf growth, we found a significant correlation (with a slope of `r tidy(lmlitup)$estimate[2]`, SE = `r tidy(lmlitup)$estimate[2]`, n = `glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig. 5).

\
We estimated gross leaf production from measured litter production ($P_L$) and temporal dynamics of $L_\tau$. Briefly, the change in $L_\tau$ was estimated for each litter collection period (again, from fitted generalized additive models to allow interpolation from a smoothed timeseries), and added to $P_L$ to arrive at an estimate of the leaf production. The temporal dynamics of litter and leaf production are shown in Fig. 5, and demonstrate the irregular temporal pattern in leaf production and $P_L$. We found no significant difference in either litter or leaf production between $C_a$ treatments (p > 0.2 for both, assessed with linear-mixed effects models).
